\titletext{}

### 2021-07-19 manuscript/regional_befr_manuscript.tex
Pre-copy-editing
Sanitised .tex

% Images 
\usepackage{graphics} % Insert images easily
\usepackage{graphicx} % Extended image support
\usepackage{float} % Graphics placement [H] [H!] arguments

\makeatletter
	\g@addto@macro\@floatboxreset\centering % Automatically centre images (floats)
\makeatother

@@

% In this study :::
In this study, we make the first known estimation of the Biodiversity-Ecosystem Function Relationship (BEFR) in disturbance-driven wooded ecosystems (savannas and woodlands), using southern Africa as our study region. We aim to understand the synergistic effects of environmental and biotic drivers of variation in the BEFR, with a view to creating a general model of the BEFR in disturbance-prone wooded ecosystems, which is currently lacking in the BEFR literature. We posit three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect of tree species diversity on woody biomass will increase with organismal density (number of stems ha\textsuperscript{-1}), with competitive interactions becoming more pronounced as stem proximity increases. Relatedly, we expect that an increase in disturbance by fire will decrease organismal density and therefore competition, weakening the effect of tree species diversity on woody biomass. Finally, we expect that (3) tree species diversity will increase tree structural diversity due to inter-specific variation in size strategy (i.e. physiognomic diversity), providing an indirect path by which tree diversity increases woody biomass.

@@

% Situation - DONE
The study used \nplots{} woodland monitoring plots from a larger pool of 5395 plots in the larger SEOSAW database \citep{SEOSAW2020}, located across 10 countries within southern Africa in the miombo ecoregion (\autoref{clust_map}, \citealp{White1987}). The study area spans the core climate space of the region, with a precipitation gradient from c. 460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to c. 1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique. A 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland ecoregion \citep{White1987}, using WorldClim estimates of Mean Annual Temperature (MAT, BIO1) and Mean Annual Precipitation (MAP, BIO12) between 1970 and 2000 with a pixel size of 30 arc seconds (926 m at equator) \citep{Fick2017}.

% Plots chosen - DONE
Plots were chosen from the SEOSAW database based on the quality and completeness of data collection, and plot configuration. Plot vegetation was identified under the broad term of `savanna', which includes `woodland', `savanna woodland', and `tree savanna', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands, or savannas \citep{Ratnam2011, Hill2010}. Plots with evidence of farming, human resource extraction, experimental treatments such as prescribed burning or herbivore exclusion, or containing termite mounds were excluded from the initial pool. Only plots >0.1 hectares were used in analyses, as area-based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate biomass estimates. Only plots with a stem density >50 trees ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots represented woodland rather than `grassy savanna', which is considered here a separate biome with very different species composition \citep{Parr2014}. \nzam{} plots within the SEOSAW database were arranged in clusters of four 20x50 m plots, with 20 m between plots. Plots within each spatial cluster were combined and treated as a single plot in analyses, resulting in \nzamcluster{} aggregate plots which were then subject to the plot filtering process described above.

@@

% AGB
We considered only trees and shrubs in our calculations of Above-Ground woody Biomass (AGB), including woody species such as palms and cycads, which are functionally tree-like. Woody lianas are scarce in our study plots and were not measured. Only living stems >10 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses. Many plots in the dataset did not include data on stems <10 cm DBH. For those plots which contained stem measurements <10 cm DBH, small stems only accounted for a median of \percsmallagb{}\% of the plot AGB.

% Number of stems
All stems >10 cm DBH were measured within each plot resulting in a total of \num[group-separator={,}]{\ntrees{}} stems with measurements. A tree may be comprised of multiple stems and so tree-level richness estimates, rather than stem-level estimates, were used to prevent bias from species which readily coppice. For each tree, we recorded species, tree height to the top of the highest branch material, and the DBH of each stem >10 cm DBH. Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks. When stem DBH could not be reliably measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m. The height of this measurement was used to estimate diameter at 1.3 m using a cubic polynomial regression to account for stem taper, with parameters estimated using a test dataset from Ryan C., (unpublished), see \citet{Godlee2020}.

% AGB Calculation
AGB for each plot (t ha\textsuperscript{-1}) was calculated as the sum of the AGB of each stem >5 cm DBH ($AGB_{i}$), divided by the plot area, using \autoref{chave_agb} taken from \citet{Chave2014}:

@@

% WorldClim, SoilGrids and MODIS
Climatic data were taken from the WorldClim database, using the BioClim variables \citep{Fick2017}. In addition to MAT and MAP, temperature stress was calculated as the mean diurnal temperature range (BIO2) and precipitation seasonality was calculated as the mean of the coefficient of variation of monthly mean precipitation (BIO15). Soil fertility data were extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot centre \citep{Hengl2017}. We extracted Cation Exchange Capacity (CEC) (cmolc kg\textsuperscript{-1}), soil organic carbon stocks (kg m\textsuperscript{-2}) percentage soil sand content (0.05-2 mm) by weight and soil nitrogen content (g kg\textsuperscript{-1}). These data are a modelled product derived from various remotely sensed and directly measured data sources. The degree of fire disturbance was calculated using the MODIS monthly burned area product at 500 m resolution (MCD64A1, \citealt{MODIS_burn}), counting the total number of times the plot pixel was classified as burning, between 2001 and 2018. We initially aimed to include disturbance by herbivory in our model, including total herbivore biomass from the \citet{Hempson2017} modelled herbivory product, but this inclusion prevented models from converging due to its collinearity with other observed variables, notably MAP and disturbance by fire.

@@

% Correlations
Pairwise correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity (extrapolated tree species richness, Shannon equitability index) and structural diversity (coefficients of variation of DBH and height) variables had moderate positive correlations with AGB (\autoref{corr_mat}, Figure S3). Stem density had the strongest correlation with AGB of all variables considered (\ccib{}). Environmental variables had weaker correlations with AGB than diversity variables, with all environmental variables having significant correlations with AGB, except fire frequency. The direction of these correlations was used as a test of our assumptions for the direction of influence of latent variables later used in the SEMs. MAP had positive correlations with all tree species diversity and structural diversity variables. Tree species diversity variables had clear positive correlations with stem density (species richness: \ccsi{}; Shannon equitability: \ccei{}), but structural diversity variables showed weak correlations with stem density (DBH CoV: \ccdvi{}, Height CoV: \cchvi{}).

@@

An anonymised version of the data that support the findings of this study are available at this DOI: .

>>>>

% Images 
\usepackage{graphics} 
\usepackage{graphicx} 
\usepackage{float}

\makeatletter
	\g@addto@macro\@floatboxreset\centering 
\makeatother

@@

In this study, we make the first known estimation of the Biodiversity-Ecosystem Function Relationship (BEFR) in disturbance-driven wooded ecosystems (savannas and woodlands), using southern Africa as our study region. We aim to understand the synergistic effects of environmental and biotic drivers of variation in the BEFR, with a view to creating a general model of the BEFR in disturbance-prone wooded ecosystems, which is currently lacking in the BEFR literature. We posit three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect of tree species diversity on woody biomass will increase with organismal density (number of stems ha\textsuperscript{-1}), with competitive interactions becoming more pronounced as stem proximity increases. Relatedly, we expect that an increase in disturbance by fire will decrease organismal density and therefore competition, weakening the effect of tree species diversity on woody biomass. Finally, we expect that (3) tree species diversity will increase tree structural diversity due to inter-specific variation in size strategy (i.e. physiognomic diversity), providing an indirect path by which tree diversity increases woody biomass.

@@

The study used \nplots{} woodland monitoring plots from a larger pool of 5395 plots in the larger SEOSAW database \citep{SEOSAW2020}, located across 10 countries within southern Africa in the miombo ecoregion (\autoref{clust_map}, \citealp{White1987}). The study area spans the core climate space of the region, with a precipitation gradient from c. 460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to c. 1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique. A 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland ecoregion \citep{White1987}, using WorldClim estimates of Mean Annual Temperature (MAT, BIO1) and Mean Annual Precipitation (MAP, BIO12) between 1970 and 2000 with a pixel size of 30 arc seconds (926 m at equator) \citep{Fick2017}.

Plots were chosen from the SEOSAW database based on the quality and completeness of data collection, and plot configuration. Plot vegetation was identified under the broad term of `savanna', which includes `woodland', `savanna woodland', and `tree savanna', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands, or savannas \citep{Ratnam2011, Hill2010}. Plots with evidence of farming, human resource extraction, experimental treatments such as prescribed burning or herbivore exclusion, or containing termite mounds were excluded from the initial pool. Only plots >0.1 hectares were used in analyses, as area-based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate biomass estimates. Only plots with a stem density >50 trees ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots represented woodland rather than `grassy savanna', which is considered here a separate biome with very different species composition \citep{Parr2014}. \nzam{} plots within the SEOSAW database were arranged in clusters of four 20x50 m plots, with 20 m between plots. Plots within each spatial cluster were combined and treated as a single plot in analyses, resulting in \nzamcluster{} aggregate plots which were then subject to the plot filtering process described above.

@@

We considered only trees and shrubs in our calculations of Above-Ground woody Biomass (AGB), including woody species such as palms and cycads, which are functionally tree-like. Woody lianas are scarce in our study plots and were not measured. Only living stems >10 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses. Many plots in the dataset did not include data on stems <10 cm DBH. For those plots which contained stem measurements <10 cm DBH, small stems only accounted for a median of \percsmallagb{}\% of the plot AGB.

All stems >10 cm DBH were measured within each plot resulting in a total of \num[group-separator={,}]{\ntrees{}} stems with measurements. A tree may be comprised of multiple stems and so tree-level richness estimates, rather than stem-level estimates, were used to prevent bias from species which readily coppice. For each tree, we recorded species, tree height to the top of the highest branch material, and the DBH of each stem >10 cm DBH. Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks. When stem DBH could not be reliably measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m. The height of this measurement was used to estimate diameter at 1.3 m using a cubic polynomial regression to account for stem taper, with parameters estimated using a test dataset from Ryan C., (unpublished), see \citet{Godlee2020}.

AGB for each plot (t ha\textsuperscript{-1}) was calculated as the sum of the AGB of each stem >5 cm DBH ($AGB_{i}$), divided by the plot area, using \autoref{chave_agb} taken from \citet{Chave2014}:

@@

Climatic data were taken from the WorldClim database, using the BioClim variables \citep{Fick2017}. In addition to MAT and MAP, temperature stress was calculated as the mean diurnal temperature range (BIO2) and precipitation seasonality was calculated as the mean of the coefficient of variation of monthly mean precipitation (BIO15). Soil fertility data were extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot centre \citep{Hengl2017}. We extracted Cation Exchange Capacity (CEC) (cmolc kg\textsuperscript{-1}), soil organic carbon stocks (kg m\textsuperscript{-2}) percentage soil sand content (0.05-2 mm) by weight and soil nitrogen content (g kg\textsuperscript{-1}). These data are a modelled product derived from various remotely sensed and directly measured data sources. The degree of fire disturbance was calculated using the MODIS monthly burned area product at 500 m resolution (MCD64A1, \citealt{MODIS_burn}), counting the total number of times the plot pixel was classified as burning, between 2001 and 2018. We initially aimed to include disturbance by herbivory in our model, including total herbivore biomass from the \citet{Hempson2017} modelled herbivory product, but this inclusion prevented models from converging due to its collinearity with other observed variables, notably MAP and disturbance by fire.

@@

Pairwise correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity (extrapolated tree species richness, Shannon equitability index) and structural diversity (coefficients of variation of DBH and height) variables had moderate positive correlations with AGB (\autoref{corr_mat}, Figure S3). Stem density had the strongest correlation with AGB of all variables considered (\ccib{}). Environmental variables had weaker correlations with AGB than diversity variables, with all environmental variables having significant correlations with AGB, except fire frequency. The direction of these correlations was used as a test of our assumptions for the direction of influence of latent variables later used in the SEMs. MAP had positive correlations with all tree species diversity and structural diversity variables. Tree species diversity variables had clear positive correlations with stem density (species richness: \ccsi{}; Shannon equitability: \ccei{}), but structural diversity variables showed weak correlations with stem density (DBH CoV: \ccdvi{}, Height CoV: \cchvi{}).

@@

An anonymised version of the data that support the findings of this study are available at this DOI: \url{https://doi.org/10.7488/ds/3095}.
